New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-09-23 paper/paper.Rmd
handled all the red/blue text

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the \textcolor{blue} {coefficients of a variable in} \textcolor{red}{contribution of a given variable to} a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the slice center, which is useful for exploring how structure changes in local subspaces. These controls are implemented in Mathematica to take advantage of \textcolor{blue}{it's} \textcolor{red}{its} \textcolor{blue}{remove this: mathematics functions, and interactivity tools} \textcolor{red}{location controls} that allow construction of functions that interact with the position of the cursor. We demonstrate the usefulness of this new tool with an example exploring region boundaries in classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approch is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many \textcolor{blue}{remove: }, many} tour developments, which are summarized in @lee2021.

@@

<!-- Purpose -->
Manual controls are especially useful for assessing sensitivity of structure to particular elements of the projection. There are many places where it is useful. In exploratory data analysis, where one sees clusters in a projection, \textcolor{red}{} one may ask whether some} variables can be removed from the projection without affecting the clustering. For interpreting models, one can reduce or increase a variable's contribution to examine the variable importance. Having the user interact with a projection is extremely valuable for understanding high-dimensional data. However, these algorithms have two problems: (1) the pre-processing of creating a manipulation space overly complicates the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to \textcolor{blue}{control}\textcolor{red}{choose} the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the value of center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. \textcolor{red}{It would also allow the user to interactively decide on the thickness of the slice.}

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods \textcolor{blue}{are}\textcolor{red}{ is} illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a mathematica notebook that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and two mathematica notebooks, one focused on the application, and another that could be used generically for new data.

@@

The problem with \textcolor{red}{the} new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalisation will change them. \textcolor{red}{Even though these changes are for the most part imperceptible, one may wish to avoid them and} there are numerous ways that this can be enforced, a few are detailed in the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}

@@

## Structure

>>>>

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the contribution of a given variable to a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the center of a slice, for example, to explore how structure changes in local subspaces. Two Mathematica notebooks are provided, which contains functions enabling the user to experiment with these new manual controls, with one specifically for exploring regions and boundaries produced by classification models. The advantage of Mathematica is its linear algebra capabilities, and interactive cursor location controls. Some limited implementation has also been made available in the R package \tt{tourr}. 
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approch is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

<!-- Purpose -->
Manual controls are especially useful for assessing sensitivity of structure to particular elements of the projection. There are many places where it is useful. In exploratory data analysis, where one sees clusters in a projection, one may ask whether some variables can be removed from the projection without affecting the clustering. For interpreting models, one can reduce or increase a variable's contribution to examine the variable importance. Having the user interact with a projection is extremely valuable for understanding high-dimensional data. However, these algorithms have two problems: (1) the pre-processing of creating a manipulation space overly complicates the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to choose the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the value of center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. It would also allow the user to interactively decide on the thickness of the slice.

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods is illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a mathematica notebook that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and two Mathematica notebooks, one focused on the application, and another that could be used generically for new data.

@@

The problem with the new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalisation will change them. Even though these changes are for the most part imperceptible, one may wish to avoid them and there are numerous ways that this can be enforced, a few are detailed in the Appendix. These primarily differ in how the remaining variables are adjusted during orthonormalisation.

@@

### 2022-09-23 paper/paper.tex
handled all the red/blue text

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the
\textcolor{blue} {coefficients of a variable in}
\textcolor{red}{contribution of a given variable to} a low-dimensional
projection, which is useful for exploring the sensitivity of structure
to particular variables. The user can also interactively shift the slice
center, which is useful for exploring how structure changes in local
subspaces. These controls are implemented in Mathematica to take
advantage of \textcolor{blue}{it's} \textcolor{red}{its}
\textcolor{blue}{remove this: mathematics functions, and interactivity tools}
\textcolor{red}{location controls} that allow construction of functions
that interact with the position of the cursor. We demonstrate the
usefulness of this new tool with an example exploring region boundaries
in classification problems.
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approch
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many \textcolor{blue}{remove: }, many\}
tour developments, which are summarized in \citet{lee2021}.

@@

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, \textcolor{red}{} one may ask whether some\}
variables can be removed from the projection without affecting the
clustering. For interpreting models, one can reduce or increase a
variable's contribution to examine the variable importance. Having the
user interact with a projection is extremely valuable for understanding
high-dimensional data. However, these algorithms have two problems: (1)
the pre-processing of creating a manipulation space overly complicates
the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to
\textcolor{blue}{control}\textcolor{red}{choose} the position of the
center of a slice. The slice tour was introduced in \citet{slicetour}.
It operates by converting the projection plane into a slice, by removing
or de-emphasizing points that are further than a fixed orthogonal
distance from the plane. The projection plane is usually thought of as
passing through the center of the data. Manual control would allow the
user to change the value of center point, by shifting it along a
coordinate axis, while keeping the orientation of the projection plane
fixed. The purpose would be to explore how or if the shape of the data,
in the space orthogonal to the projection, changes as one gets away from
the center.
\textcolor{red}{It would also allow the user to interactively decide on the thickness of the slice.}

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods
\textcolor{blue}{are}\textcolor{red}{ is} illustrated to compare and
contrast boundaries constructed by different classifiers. The software
section describes a mathematica notebook that is used for the
application, and describes the interactive environment that would be
desirable within R as new technology becomes available. The paper is
accompanied by an appendix with more details and adjustments to the
manual controls, and two mathematica notebooks, one focused on the
application, and another that could be used generically for new data.

@@

The problem with \textcolor{red}{the} new simple method is that it is
not faithful to the precise values for \(V_m\) because the
orthonormalisation will change them.
\textcolor{red}{Even though these changes are for the most part imperceptible, one may wish to avoid them and}
there are numerous ways that this can be enforced, a few are detailed in
the Appendix.
\textcolor{red}{While they all enforce the same position for the selected variable, they differ in how the remaining variables are adjusted during orthonormalisation.}

@@

\hypertarget{structure}{%
\subsection{Structure}\label{structure}}

>>>>

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the contribution of a given variable to a
low-dimensional projection, which is useful for exploring the
sensitivity of structure to particular variables. The user can also
interactively shift the center of a slice, for example, to explore how
structure changes in local subspaces. Two Mathematica notebooks are
provided, which contains functions enabling the user to experiment with
these new manual controls, with one specifically for exploring regions
and boundaries produced by classification models. The advantage of
Mathematica is its linear algebra capabilities, and interactive cursor
location controls. Some limited implementation has also been made
available in the R package \tt{tourr}. 
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approch
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many tour developments, which are
summarized in \citet{lee2021}.

@@

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, one may ask whether some variables can be
removed from the projection without affecting the clustering. For
interpreting models, one can reduce or increase a variable's
contribution to examine the variable importance. Having the user
interact with a projection is extremely valuable for understanding
high-dimensional data. However, these algorithms have two problems: (1)
the pre-processing of creating a manipulation space overly complicates
the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to
choose the position of the center of a slice. The slice tour was
introduced in \citet{slicetour}. It operates by converting the
projection plane into a slice, by removing or de-emphasizing points that
are further than a fixed orthogonal distance from the plane. The
projection plane is usually thought of as passing through the center of
the data. Manual control would allow the user to change the value of
center point, by shifting it along a coordinate axis, while keeping the
orientation of the projection plane fixed. The purpose would be to
explore how or if the shape of the data, in the space orthogonal to the
projection, changes as one gets away from the center. It would also
allow the user to interactively decide on the thickness of the slice.

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods is illustrated
to compare and contrast boundaries constructed by different classifiers.
The software section describes a mathematica notebook that is used for
the application, and describes the interactive environment that would be
desirable within R as new technology becomes available. The paper is
accompanied by an appendix with more details and adjustments to the
manual controls, and two Mathematica notebooks, one focused on the
application, and another that could be used generically for new data.

@@

The problem with the new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalisation will change
them. Even though these changes are for the most part imperceptible, one
may wish to avoid them and there are numerous ways that this can be
enforced, a few are detailed in the Appendix. These primarily differ in
how the remaining variables are adjusted during orthonormalisation.

@@
